Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-02-10 paper/main.tex
Adds notation

\section{Background}
\label{sec:background}
\todo[inline]{Need to rewrite this}
We denote random variables with uppercase letters $ X $, and a set of random
variables as $ \bm{X} = \{X_1, \cdots, X_k\} $ with $ \rvert \bm{X} \rvert = k
$. A sample from the random variable $ X $ is denoted as $ x $ and from a set
of random variables $ \bm{X} $ is denoted as $ \bm{x} $. In this paper, we
consider random variables in the mixed data setting where each of the random
variables can be either of continuous, categorical, or ordinal unless
specified. We write the expectation of a variable $ X $ as $ \mathbb{E}[X] $,
conditional expectations as $ \mathbb{E}[X | \bm{Z}] $. We denote a graph $ G =
(V, E) $ which nodes $ V$ and edges $ E $. In this paper we focus on all
observed variables, and linear measure of association. 
where $ \mathrm{cov}(X, Y) $ is the covariance between $ X $ and $ Y $, $ \mathrm{corr}(X, Y) $ is the correlation between $ X $ and $ Y $, and $
\sigma_X $ and $ \sigma_Y $ are the standard deviations of $ X $ and $ Y $
respectively.  Covariance matrix using $ \Sigma $ and the entry corresponding to
covariance between $ X $ and $ Y $ in the matrix using $ \Sigma_{XY} $.

\subsection{Model Testing Using CI Tests}
Our structure learning algorithm relies on the idea of testing DAGs using CI
tests. In this section, we give an example of how this model testing and
corresponding modification work. We take the example of the MEC learned by the
PC algorithm in Figure~\ref{fig:intro}(b) using $ N=800 $. We first orient all
the undirected edges in the model based on our domain knowledge. In this case
we only have one undirected edge between Marital Status and Relationship. Let's 
orient the edge as Marital Status to Relationship.

 The implied CIs of the 
model along with the test results are shown in the Figure.

We can think of our proposed approach to be an iterative application of this model
testing approach.

@@

Our structure learning approach relies on using CI tests and a measure of
conditional association between variables -- also known as partial association.
Specifically, we are interested in testing CI tests: $ X \ci Y \bm{Z} $, i.e.,
whether the variables $ X $ and $ Y $ are independent when conditioned on $
\bm{Z} $. Secondly, we are interested in quantifying the strength of
relationship between $ X $ and $ Y $ conditioned on $ \bm{Z} $. We can use
the effect size measures of the CI tests for this quantification. In this 
section, we give an overview of some of the effect size measures for 
different types of (possibly mixed) data.

@@

\paragraph{$ X $ is ordinal, and $ Y $ and $ Z $  are continuous or ordinal: }

@@

\paragraph{$ X $, $ Y $, and $ \bm{Z} $ are discrete (ordinal or caterical): }

>>>>

\section{Background}
\label{sec:background}
We denote random variables using uppercase letters, such as $ X $, and a set of
random variables by $ \bm{X} = \{X_1, \cdots, X_k\} $, where $ \rvert \bm{X}
\rvert = k $ indicates the number of variables in the set. We denote the
covariance between variables $ X $ and $ Y $ as $ \mathrm{cov}(X, Y) $, the
covariance matrix as $ \Sigma $, and the entry corresponding to covariance
between $ X $ and $ Y $ using $ \Sigma_{XY} $. In the context of this paper, we
consider causal DAGs that contain only observed variables. The structure of the
DAG is represented using a graph $ G = (V, E) $, where $ V $ are the nodes
corresponding to the random variables, and $ E $ are the edges denoting the
causal relationship between the variables. The set of parents of $ X $ in $ G $
are denoted as $ \textrm{Pa}_G(X) $, and its ancestors and descendants as $
\textrm{An}_G(X) $ and $ \textrm{De}_G(X) $ respectively. We denote a path, $
\pi(X, Y) = \{ X, V_0, V_1, \cdots, V_k, Y \} $ between $ X $ and $ Y $ in $ G
$ where each consecutive pair of variables in $ \pi $ are connected through an
edge. We say $ X $ and $ Y $ are d-connected in $ G $ given a conditioning set
$ \bm{Z} \subseteq V - \{X, Y\} $, if for a path $ \pi(X, Y) $ (i) every collider 
structure ($ V_1 \rightarrow V_2 \leftarrow V_3 $) on $ \pi(X, Y) $, $ \bm{Z} \cap \{ V_2, \textrm{De}_G(V_2) \} \ne \emptyset $, and (ii) for every non-collider structure
($ V_1 \rightarrow V_2 \rightarrow V_3 $, $ V_1 \leftarrow V_2 \rightarrow V_3 $), $ V_2 \not \in \bm{Z} $. $ X $ and $ Y $ are d-separated given $ G $ if no d-connecting
path exists between them.

\subsection{Testing DAGs Using CI Tests}
Our structure learning algorithm uses the idea of testing DAGs using CI tests.
In this section, we give an example of how this model testing works in
practice. We take the example of the MEC learned by the PC algorithm in
Figure~\ref{fig:intro}(b) using $ N=800 $. We first orient all the undirected
edges in the model based on our domain knowledge to obtain a DAG. In this case
we only have one undirected edge between Marital Status and Relationship. Let's
orient the edge as Marital Status to Relationship.

Using these statistical tests we can determine the pair of variables that
should have edge between them. We can think of our proposed approach to be an
iterative application of this model testing approach.

@@

Our structure learning approach relies on using CI tests and a measure of
conditional association between variables -- also known as partial association.
Specifically, we are interested in testing CI tests: $ X \ci Y \rvert \bm{Z} $
(with potentially $ \bm{Z} = \emptyset $. Secondly, we are interested in
quantifying the strength of the relationship between $ X $ and $ Y $
conditioned on $ \bm{Z} $. We can use the effect size measures of the CI tests
for this quantification. In this section, we give an overview of some of the
effect size measures for different types of (possibly mixed) data.

@@

\paragraph{$ X $ is ordinal, and $ Y $ and $ \bm{Z} $  are continuous or ordinal: }

@@

\paragraph{All $ X $, $ Y $, and $ \bm{Z} $ are discrete (ordinal or categorical): }

In the case when we have a combination of ordinal and categorical variables, we use
the
